Supplemental Information: Aligning marine species range data to better serve science and conservation

### 2016-09-01 ms_3_res_disc_concl.Rmd
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

[results] The two datasets have notably different taxonomic (Fig 1A) and regional (Figs 1B, 1C) coverage.  AquaMaps encompasses a broader range of taxa than IUCN, as IUCN spatial data files are only available for select taxonomic groups that have been comprehensively assessed. While species numbers in both datasets peak in tropical latitudes near the equator, species counts for IUCN maps drop quickly beyond 30°N and 30°S, while species counts for AquaMaps remain robust well into temperate latitudes.  The longitude frequency plots show a slight shift in the IUCN dataset away from the Atlantic and eastern Pacific compared to AquaMaps.

@@

[discussion] To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (how much of the smaller range falls within the larger range, i.e., where on the map) against the area ratio (ratio of smaller range area to larger range area, i.e., how much of the map) for each shared species (Fig 2A).  Where two independent methods predict presence (or absence) of a species in the same location, we can assume a higher confidence in the accuracy of that prediction.  Where one method predicts presence while the other predicts absence, we can investigate potential sources of commission and omission errors to refine our understanding of the species' range.  Understanding these sources of error can provide guidance to improve predictions of species range even when only one source of spatial data is available.

@@

[results] In constraining coral ranges to shallow waters, we see a strong increase in the apparent alignment of species maps between IUCN and AquaMaps.  Using the original quadrant definitions from Fig 2B, we see in Fig 3B a massive shift in 354 paired-map coral species due to constraining coral depths to 200 m.  Membership in the "well-aligned" quadrant jumped from 22.4% to 76.2%, with a corresponding decrease in all other quadrants.  See SXXX to examine the shifts of individual species among the quadrants.

>>>>

[results] The two datasets have notably different taxonomic (Fig 1A) and regional (Figs 1B, 1C) coverage.  AquaMaps encompasses a broader range of taxa than IUCN, as IUCN spatial data files are only available for select taxonomic groups that have been comprehensively assessed. While species numbers in both datasets peak in tropical latitudes near the equator, species counts for IUCN maps drop quickly beyond 30°N and 30°S, while AquaMaps includes distribution of species well into temperate latitudes.  The longitude frequency plots show a slight shift in the IUCN dataset away from the Atlantic and eastern Pacific compared to AquaMaps.

@@

[discussion] To explore differences in species distribution and range between the two datasets, we assessed the distribution alignment (how much of the smaller range falls within the larger range, i.e., where on the map) relative to the area ratio (ratio of smaller range area to larger range area, i.e., how much of the map) for each shared species (Fig 2A).  Where two independent methods predict presence (or absence) of a species in the same location, we can assume a higher confidence in the accuracy of that prediction.  Where one method predicts presence while the other predicts absence, we can investigate potential sources of commission and omission errors to refine our understanding of the species' range.  Understanding these sources of error can provide guidance to improve predictions of species range even when only one source of spatial data is available.

@@

[results] In constraining coral ranges to shallow waters, we see a strong increase in the apparent alignment of species maps between IUCN and AquaMaps.  Using the original quadrant definitions from Fig 2B, we see in Fig 3B a massive shift in 354 paired-map coral species due to constraining coral depths to 200 m.  Membership in the "well-aligned" quadrant jumped from 22.4% to 76.2%, with a corresponding decrease in all other quadrants.  See Fig SXXX to examine the shifts of individual species among the quadrants.

### 2016-09-01 ms_6_supplemental.Rmd
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

from PNAS http://www.pnas.org/site/authors/preparation.xhtml:

The main text of the paper must stand on its own without the SI. Refer to SI in the manuscript at an appropriate point in the text. Number supporting figures and tables starting with S1, S2, etc. Authors are limited to no more than 10 SI files, not including movie files.

Authors who place detailed materials and methods in SI must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the online methods. If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.

Because PNAS edits SI and composes it into a single PDF, authors must provide the following file formats only:

Text: Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
Figures: Provide a brief legend for each supporting figure after the supporting text. Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or GIF format; figures may not be embedded in manuscript text. When saving TIFF files, use only LZW compression; do not use JPEG compression. Do not save figure numbers, legends, or author names as part of the image. Composite figures must be preassembled.
3D figures: Supply a composable U3D or PRC file so that it may be edited and composed. Authors may submit a PDF file but please note it will be published in raw format and will not be edited or composed.
Tables: Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file. Do not use tabs or spaces to separate columns in Word tables.
Datasets: Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
Movies: Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files and submit a brief legend for each movie in a Word or RTF file. All movies should be submitted at the desired reproduction size and length. Movies should be no more than 10 MB in size.
Still images: Authors must provide a still image from each video file. Supply TIFF, EPS, high-resolution PDF, JPEG, or GIF files.
Appendices: PNAS prefers that authors submit individual source files to ensure readability. If this is not possible, supply a single PDF file that contains all of the SI associated with the paper. This file type will be published in raw format and will not be edited or composed.
-->
<!--
# Data sources and methods

All code is publicly available and freely downloadable [here on GitHub](link to github).

## AquaMaps

AquaMaps data was provided by AquaMaps directly via ftp. Three SQL files were sent that contained (1) list of all 22,889 modeled species and their taxonomic information, (2) information for all (259,200) half degree cells used for range mapping and (3) the probability of occurrence per cell for all 22,889 species. These were extracted into .csv form using R Statistical Software.

## IUCN

As of December 2015, IUCN had published species distribution maps for XXX marine species across 24 taxonomic groups.  For this analysis, we did not consider IUCN range maps for bird species, as those data are hosted separately by BirdLife International.

Each shapefile contains spatial polygons outlining limits of distribution for each species, indicating regions of presence/absence; additional attributes provide information on extant/extinct ranges, native/introduced ranges, and seasonality.
-->

@@

![](figures/si_four_maps.png)

>  __Fig. S1.__ Representative species maps to illustrate each quadrant from Fig. 2A.  Note that B-E are arranged to match quadrants in Fig. S1A. (A) Alternative version of Figure 2A that includes extinction risk categories for each species.  Four representative species, one from each quadrant, are mapped in (B-E).
(B) Well-matched: Kajikia albida, the Atlantic white marlin.  Ranges from each data set show nearly complete overlap, and very similar area of coverage.
(C) Distribution-matched: Conus episcopatus, the dignified cone snail.  Ranges shows excellent overlap in the western Pacific, though IUCN range extends beyond the bounds of the AquaMaps range.
(D) Area-matched: Conus magnificus, the magnificent cone snail.  Ranges overlap in the southern Pacific, but align poorly elsewhere.  The total area for each range is similar.
(E) Poorly-matched: Acanthurus nigroris, the blue-lined surgeonfish.  IUCN predicts species range only near the Hawaiian islands; AquaMaps predicts extensive range throughout the central and western Pacific Ocean.  The datasets align in neither distribution nor area of coverage.

@@

>  __Fig. S2.__ Breakdown of quadrants (Fig. S1A) by IUCN extinction risk categories.  IUCN and Aquamaps ranges were in better agreement (i.e., had a higher probability of being in the upper-right, "well-aligned" quadrant, Figure 2A/Figure S1A) for species at greater risk according to IUCN risk categories (p < 0.001, generalized linear model, family = binomial, Table S1).  This suggests at-risk species may receive more attention in terms of data collection or model refinement.

@@

>  __Fig. S3.__ IUCN-defined range of _Oculina varicosa_, a typical photosynthetic coral, plotted against a 200 m bathymetry contour. According to IUCN, _O. varicosa_ can be found to depths of 152 m.  Maps of many other coral species and other reef-associated organisms followed similar patterns.

@@

>  __Fig. S4.__ AquaMaps distribution map extent remaining after applying a presence threshold. A higher threshold constrains an analysis to cells with near certainty of occurrence, while a low threshold captures larger areas of increasingly marginal suitability. (A) A 40% threshold applied to all species in the AquaMaps dataset shows a mean loss of 28% of total range compared to a 0% threshold, with a wide distribution in which some species lose nearly all of their apparent range.  (B) Median remaining extent at increments of presence threshold. Dark green ribbon includes 25% to 75% quantiles of remaining species range, while lighter ribbon includes 5% to 95% quantiles.  Incrementing the presence threshold from 0.00 to 1.00 for the entire AquaMaps dataset, the shallow downward trend indicates a low but consistent sensitivity of apparent species range to threshold choice, with no surprising tradeoffs that could suggest an "optimal" threshold.

>>>>

-->

methods:
IUCN raw -> grid -> raster -> final
AM raw -> presence
bathymetry poly -> raster?

results/discussion:
quad plot with risk and four species (?)
four representative maps

coral quadrant shifts

AM thresholds? not mentioned in body of paper as of yet

## Preparing AquaMaps and IUCN data layers

__Fig S1.__  Examples of AquaMaps and IUCN raw species range data. (A) AquaMaps distribution of _Thunnus alalunga_ (Albacore Tuna) showing global range with varying probabilities of occurrence assigned to 0.5° grid cells. (B) IUCN distribution of _T. alalunga_ represented as extent of occurrence polygons.

__Fig S2.__ Rasterizing shapefiles provided by IUCN. A portion of the _T. alalunga_ range map is used to exemplify the rasterization process.  To enable direct comparison of IUCN species ranges to AquaMaps species ranges, the raw IUCN polygon (A) is overlaid with a 0.5° degree grid matching the AquaMaps grid (B).  Each cell is assigned a value of "present" if the cell overlaps any portion of the polygon (C).  The resulting raster (D).

@@

__Fig S3.__ Representative species maps to illustrate each quadrant from Fig 2A.  Each map is positioned to match its quadrant in Fig 2A.
(A) Distribution-matched: Conus episcopatus, the dignified cone snail.  Ranges shows excellent overlap in the western Pacific, though IUCN range extends beyond the bounds of the AquaMaps range.
(B) Well-matched: Kajikia albida, the Atlantic white marlin.  Ranges from each data set show nearly complete overlap, and very similar area of coverage.
(C) Poorly-matched: Acanthurus nigroris, the blue-lined surgeonfish.  IUCN predicts species range only near the Hawaiian islands; AquaMaps predicts extensive range throughout the central and western Pacific Ocean.  The datasets align in neither distribution nor area of coverage.
(D) Area-matched: Conus magnificus, the magnificent cone snail.  Ranges overlap in the southern Pacific, but align poorly elsewhere.  The total area for each range is similar.

@@

>  __Fig S2.__ Breakdown of quadrants (Fig S1A) by IUCN extinction risk categories.  IUCN and Aquamaps ranges were in better agreement (i.e., had a higher probability of being in the upper-right, "well-aligned" quadrant, Figure 2A/Figure S1A) for species at greater risk according to IUCN risk categories (p < 0.001, generalized linear model, family = binomial, Table S1).  This suggests at-risk species may receive more attention in terms of data collection or model refinement.

@@

>  __Fig S3.__ IUCN-defined range of _Oculina varicosa_, a typical photosynthetic coral, plotted against a 200 m bathymetry contour. According to IUCN, _O. varicosa_ can be found to depths of 152 m.  Maps of many other coral species and other reef-associated organisms followed similar patterns.

@@

>  __Fig S4.__ AquaMaps distribution map extent remaining after applying a presence threshold. A higher threshold constrains an analysis to cells with near certainty of occurrence, while a low threshold captures larger areas of increasingly marginal suitability. (A) A 40% threshold applied to all species in the AquaMaps dataset shows a mean loss of 28% of total range compared to a 0% threshold, with a wide distribution in which some species lose nearly all of their apparent range.  (B) Median remaining extent at increments of presence threshold. Dark green ribbon includes 25% to 75% quantiles of remaining species range, while lighter ribbon includes 5% to 95% quantiles.  Incrementing the presence threshold from 0.00 to 1.00 for the entire AquaMaps dataset, the shallow downward trend indicates a low but consistent sensitivity of apparent species range to threshold choice, with no surprising tradeoffs that could suggest an "optimal" threshold.
